---
title: 'Capstone: Exploratory Data Analysis'
author: "Shyamala Gurumurthy"
date: "March 6, 2018"
output: 
  html_document: 
    highlight: tango
    keep_md: yes
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

```{r load_libraries, warning=FALSE, message=FALSE, results='hide'}
library(tidyverse)
library(leaflet)
```

### Reduce number of categories 
Number of categories were reduced from 39 to 15 by combining common crime types under one category. This reduction in detail is to help gain quick insights while plotting the data. It is not expected to significantly affect the analysis.

```{r reduce_categories, warning=FALSE, message=FALSE, results='hide'}
Categories <- read_csv("Categories.csv")
```
### `incidents_house_price` dataset
Load the `incidents_house_price` dataset from the previous data cleaning exercise.
```{r incidents_house_price, warning=FALSE, message=FALSE, results='hide'}
incidents_house_price <- read_csv("incidents_house_price.csv")
```

### Reduce categories in `incidents_house_price` dataset
```{r incidents_new_categories, warning=FALSE, message=FALSE, results='hide'}
incidents_new_categories <- left_join(incidents_house_price, Categories, by = "Category") %>%
  select(-n) %>%
  rename(New_Category = `New Category`) 
```

### Extract only features useful for plotting on a map
```{r zipc, warning=FALSE, message=FALSE, results='hide'}
zipc <- incidents_new_categories %>% 
  select(zipcode, New_Category, latitude, longitude) %>%
  count(zipcode, New_Category, latitude, longitude) %>%
  arrange(desc(n))
```

```{r filt_zipc, warning=FALSE, message=FALSE, results='hide'}
filt_zipc <- zipc %>% filter(n > 10) %>%
  mutate(leaflet_labels = paste0(New_Category, " (", n, ")")) %>%
  mutate(leaflet_radius = findInterval(n, c(50, 100, 200, 300, 400, 500, 600, 
                                            700, 800, 900, 1000, 2000, 5000, 10000)))

filt_zipc$New_Category <- as.factor(filt_zipc$New_Category)
```


### Number of crimes by category
The number of crimes for each category are shown on a map of San Francisco using the `leaflet` library. This provides a spatial visualization of areas with high concentration of crimes and crime-types. The tool tip indicates the category followed by the number of crimes in parenthesis `()`. 
```{r plot_map_filt_zipc}
col_pal <- colorFactor(palette = "magma", levels = 
                         levels(filt_zipc$New_Category))

leaflet() %>%
  setView(lng = -122.4164, lat = 37.7766, zoom = 12) %>%
  addTiles() %>%
  addCircleMarkers(filt_zipc, lng = filt_zipc$longitude, lat = filt_zipc$latitude, 
                   weight = 5, radius = filt_zipc$leaflet_radius * 1.5, fillOpacity = 0.8,
                   color = col_pal(filt_zipc$New_Category),
                   label = filt_zipc$leaflet_labels) %>%
  addLegend("topright", col_pal, values =  filt_zipc$New_Category)
```
### Number of crimes for each category
The bar plot below provides a relative comparison of the number of crimes for each category.
```{r plot_ncrimes}
incidents_new_categories  %>%
  ggplot(aes(x = New_Category)) +
  geom_bar() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_y_continuous(labels=function(n){format(n, scientific = FALSE)},
                     breaks = seq(0,500000,by = 25000)) + 
  labs(x = "Categories of crime", y = "Number of crimes") + 
  coord_flip()
```

### Incidents by day
The pie chart below is to assess the variation in the number of crimes by day of the week.
```{r incidents_by_day}
incidents_by_day <- incidents_new_categories %>%
                    select(DayOfWeek, New_Category) %>%
                    count(DayOfWeek) 
```

```{r pie_ncrimes}
pie <- incidents_by_day %>% 
       ggplot(aes(x = factor(1), y = n, fill = DayOfWeek)) + 
       geom_bar(width = 1,stat="identity", color = "black") + 
       guides(fill=guide_legend(override.aes=list(colour=NA))) +
       coord_polar(theta = "y") + 
       theme(axis.ticks=element_blank(),  
             axis.title=element_blank(),  
             axis.text.y=element_blank()) 

y.breaks <- cumsum(incidents_by_day$n) - incidents_by_day$n/2

pie + 
  theme(axis.text.x=element_text(color='black')) +
  scale_y_continuous(breaks=y.breaks, labels=incidents_by_day$DayOfWeek) + 
  geom_text(aes(y=y.breaks, label = n), size=3)
```

### Incidents by month and year
The heatmap below shows the number of crimes by month and year. It allows for easy visualization of hot-spots, i.e., month-year combinations with high crime-rates.

```{r incidents_by_month}
incidents_by_month <- incidents_new_categories %>%
                     select(year, month) %>%
                     count(year, month)
```

```{r plot_incidents_by_month}
incidents_by_month %>%
  mutate_at(c("month"), as.factor) %>%
  mutate_at(c("year"), as.factor) %>%
  filter(n > 7000) %>%
  ggplot(aes(x = month, y = year)) +
    geom_tile(aes(fill = n)) + 
    scale_fill_gradient(low = "white", high = "darkred") +
    labs(x = "Months", y = "Number of crimes") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0)) + 
    scale_x_discrete(labels=month.name) 
```

### Number of incidents resolved by each PD district
The bar plot below shows the number of incidents resolved for each PD district. The visualization allows for easy comparison of the relative resolution of the number of cases for each district.
```{r resolution}
resolution <- incidents_new_categories %>%
  select(New_Category, PdDistrict,Resolution) %>%
  mutate(Resolved = ifelse(Resolution == "NONE" , 0, 1))
```

```{r plot_resolution, fig.width=12, fig.height=6}
resolution %>%
  mutate(Resolved = ifelse(Resolved == 0, "N", "Y")) %>%
  mutate_at(c("Resolved"), .funs = as.factor) %>%
  ggplot(aes(x = Resolved)) + 
  geom_bar(aes(color = Resolved, fill = Resolved)) + 
  facet_grid(~PdDistrict) +
  scale_fill_manual(values = c("DarkRed", "DarkGreen")) +
  scale_color_manual(values = c("DarkRed", "DarkGreen"))  +
  theme_bw() +
  theme(legend.position="none")
```